---
title: "Practical Machine Learning Project"
author: "Moustafa A. Mahmoud"
date: "March 24, 2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

#Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

#Data

Data
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

#Project Goal

The goal of your project is to predict the manner in which they did the exercise. This is the **classe** variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

#Loading the Dataset
In this section, we will start loading the dataset and the required libraries.

##Load the required libraries
```{r loadlib,message=FALSE, warning=FALSE}
library(caret)
library(rpart)
library(randomForest)
```

We will crawl the dataset from link provided in Data section.

```{r loaddata}
# setting working directory
setwd("~/Master/CombinedWorkspace/R/7_Practical_Machine_Learning")

# Download the training data
download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
              destfile = "./pml-training.csv", method = "curl")
# Load the training dataset
trainData <- read.csv("./pml-training.csv", na.strings=c("NA","#DIV/0!",""))
# Download the testing data
download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
              destfile = "./pml-testing.csv", method = "curl")
# Load the testing dataset
testData <- read.csv("./pml-testing.csv", na.strings=c("NA","#DIV/0!",""))
```

##Data Analysis
In this subsection, we will do some analysis on the training dataset.
```{r analysis}
head(trainData, 10)
#str(trainData)#commented to reduce spaces
table(trainData$classe)
prop.table(table(trainData$classe))

```

#Data Cleansing

In this section, we will apply the data cleansing rules based on our data analysis task.

##Compare Train and Test Data columns
check if there are some columns available in train and not in test data (except classe)
```{r cleansing}
trainCol = colnames(trainData)
testCol = colnames(testData)
print(trainCol[!(trainCol %in% testCol)]) # only classe 

```
##Cleaning the dataset.
The first 7 columns contains ids, user names, and some timestamp data. These columns are not important to solve our problem.

```{r remove_not_used_features}
importantFeaturesNames <- colnames(trainData)[8:59]
print(importantFeaturesNames)

trainData <- trainData[,c(importantFeaturesNames,"classe")]
testData <- testData[,c(importantFeaturesNames)]

```

Also, we will remove all columns that contains at least 90% of NA or blank values on the training dataset

```{r}
# Here we get the indexes of the columns having at least 90% of NA or blank values on the training dataset
indColWithNull <- which(colSums(is.na(trainData) |trainData=="")>0.9*dim(trainData)[1]) 
trainDataCleaned <- trainData[,-indColWithNull]
testDataCleaned <- testData[,-indColWithNull]


```
#Split The Dataset

In this part, we will split the training dataset into 60% training and 40% testing.

```{r split}
set.seed(9999)
inTrain <- createDataPartition(trainDataCleaned$classe, p=0.6, list=FALSE)
training <- trainDataCleaned[inTrain,]
testing <- trainDataCleaned[-inTrain,]
```

#Fit Random Forest on the training dataset

We will fit random forest model in our training dataset. 

```{r train}
set.seed(9999)
trControl <- trainControl(method="cv", number=3, verboseIter=FALSE)
modelRF <- train(classe~., data=training, method="rf", trControl=trControl, verbose=FALSE)
modelRF$finalModel

```
#Test the model
We will test our model based on our 40% sample data splitted before.

```{r test}
predictRF <- predict(modelRF, testing)
cmRF <- confusionMatrix(predictRF, testing$classe)
cmRF

```
The accuracy rate using the random forest is very high: **Accuracy : 0.9648228** and **out of sample error is equal to 0.03517716**.

```{r}
# true accuracy of the predicted model
predicted <- predict(modelRF, testing)
outOfSampleErrorAccuracy <- sum(predicted == testing$classe)/length(predicted)
print(outOfSampleErrorAccuracy)

outOfSampleError <- 1 - outOfSampleErrorAccuracy
print(outOfSampleError)

```
#Model Validation
In this section, we will validate our model based on the testing data provided in the project.

```{r}
predictedRF <- predict(modelRF, testDataCleaned)
predictedRF
```
